GPU-Beschleunigung der Material Point Method

### 2019-01-28 mscthesis.tex
moore out
first review done

\noindent Countingsort für jede Partikel-Variable erhöht Coalescing und L2-Cache Hitraten. Binning teilt das Gitter in Blöcke auf, die Shared-Memory Implementationen ermög\-lich\-en. Die Größen der Bins sind für keine Operation beschränkt. Weiterhin werden nur Blöcke ausgeführt, die Informationen erhalten.

\vfill
\begin{center}
\subsubsection*{Abstract}
\end{center}
The Material Point Method is allowing for physically based simulations. It has found its way into computer graphics and since then rapidly expanded. The Material Point Method's hybrid use of Lagrangian particles as a persistent storage and a background uniform Eulerian grid enable solving of various partial differential equations with ease.

\noindent The Material Point Method suffers from high execution times and is thus only viable for hero shots. The method is however highly parallelizable. Thus this thesis proposes how to accelerate the Material Point Method using GPGPU techniques.

\noindent Core of the Material Point Method are grid and particles transfers that interpolate between the two structures. These transfers are executed multiple times per physical time step. Preprocessing steps might be taken if their computing time is outweighed.

\noindent Deep sorting with counting sort increases coalescing and L2 Cache hit rates. Binning allows to divide the grid into blocks for shared memory filtering techniques. All operations do not rely on fixed bin size. As another preprocessing step only grid blocks are executed which have particles in them.
\vfill
\clearpage
\tableofcontents

@@

\section{Introduction} \label{intro}
General Purpose Computation on Graphics Processing Unit(GPGPU) has elevated Graphics Processing Units(GPUs) to any computational field utilizing massively parallelizable algorithms. As such is the nature of discretizations in physics. GPU's performance still follows Moore's Law. In unison with efficient algorithms they still have much potential to offer.

The higher parallelization acceleration of GPUs is a blessing for those applications that can profit from it. Conversely algorithms which are not parallelizable perform much worse and should be avoided. Dividing up these tasks between CPU and GPU is not recommendable. Transferring data over the PCI-Bus between the two is slow for big amounts of data and requires synchronization. Tasks in such a pipeline should thus stay on the GPU although they might perform worse.

@@

The simulation of the dynamics of materials is still extremely challenging in different applications and fields. Multiphysics interactions between materials are not realized on a big scale and often are not possible or feasible. The Material Point Method has already reliable tackled many different models, materials, and even interaction between materials.

Pure Eulerian methods like the Finite-Element-Method(FEM) have been in use for long in engineering tasks but face problems handling topology changes with numerical stability. Pure Langrangian methods form the end other of the spectrum. Prominent are Smoothed Particle Hydrodynamics(SPH) that are restricted to (viscous) fluids and are of empirical nature. Particle-in-Cell(PIC) techniques like the Material Point Method are hybrid Lagrangian/Eulerian methods. They try to benefit from the nature of both. Advection on particles is trivial. Thus the Material Point Method utilizes Lagrangian particles as a consistent storage of dynamic properties. The governing equations of the physical problem can be discretized easily to and solved on a uniform Eulerian grid. The grid is unmoving and does not store any information between physical time steps.

Transfers between the Lagrangian and Eulerian are of need to bring the two together. These transfers are the focus of the Particle-in-Cell techniques. These transfers are realized due to interpolation. Thus they create (estimable) numerical error and lead to dissipation. Moreover grid nodes are mostly outnumbered by particles leading to more numerical error as modes of the single particles are not realized by the grid.

The $p$ particles to one node relationship of a Particle-to-Grid(P2G)-Transfer or inversely one node to $p$ particles of a Grid-to-Particle(G2P)-Transfer is the main problem when realizing these operations on the GPU. When faced with the static architecture of the GPU, this will inevitably lead to branch divergence of threads within a subgroup as some nodes are assigned more particles than others.

Chapter \ref{sec:rel_work} looks at the relevant literature of PIC-Methods specifically on the GPU as well as Material Point Method extensions. In chapter \ref{sec:notation} notations which will be used throughout the thesis are laid out. The Basics of chapter \ref{sec:basics} cover the derivation of an elastic material in the MPM to get an understanding of the prominent operations. Furthermore GPU intrinsics, metrics and optimization opportunities are discussed. Chapter \ref{sec:implementation} tends to the implementation. The evaluation of the implementation follows up in chapter \ref{sec:eval}. As a conclusion chapter \ref{sec:conclusion} gives a summarizing overview and tends to future work.

\clearpage
\section{Related Work}\label{sec:rel_work}
\textbf{Particle-in-Cell} techniques were developed initially for hydrodynamics \cite{evans1957particle}. They are also widely used in plasma simulations to solve Maxwell's equations \cite{PIC:GPU}. Their key steps are identical to the MPM:
\begin{enumerate}
  \item Transfer particle data to a grid.
\item \label{it:solve_gov_eq}Solve the governing partial differential equations on the grid to move forward in time.
  \item Transfer grid data back to particles.
\end{enumerate}
In general the grid is regarded as a scratch-pad since its only a temporary storage for the means of computing item \ref{it:solve_gov_eq}. I.e. the grid could be deleted each physical times step. Based on PIC to reduce dampening and momentum conservation Fluid-Implicit-Particle(FLIP) was developed for fluid simulations and is still widely in use for large simulations. Their counterpart being the empirical Smoothed Particle Hydrodynamics \cite{gingold1977smoothed}.

As a further development the Affine-Particle-in-Cell(APIC) technique was introduced as an improvement to PIC and FLIP solvers. Due to the definition of a local affine velocity field around each particle (instead of a single vector) APIC conserves angular momentum, reduces damping and improves stability over PIC and FLIP, more in chapter \ref{sec:apic}. \cite{MPM:APIC} \cite{MPM:OLD_APIC}

@@

The \textbf{Material Point Method} is a further development to the PIC methods. It uses analysis and numerics to be derived without the need for empirical assumptions. Originally it was developed by \cite{sulsky1995application} with a focus on the dynamics of solids. \cite{MPM:SNOW} brought the method into computer graphics for their animation film \textit{Frozen}. In rapid succession a variety materials and effects got modeled by the MPM. Since MPM is based on discretizing governing equations and using constitutive models covered by the different literature, its fast rate of expenditure is not a surprise.

@@

There exist several implementations utilizing the \textbf{GPU for Particle-in-Cell} techniques, which the Material Point Method is a part of:

\textbf{Particle sorting:} \cite{PIC:GPU} studies different particle list sorting methods. Methods like 'Message Passing Sort' \cite{kong2011particle} \cite{decyk2011adaptable}, 'In Place Particle-QuickSort' \cite{stantchev2008fast} and 'Linked List Ordering' \cite{burau2010picongpu} try to utilize the partial sorting and only update particles that need to be moved within the list. These methods often rely on the assumption that only a small number of particles change their grid node, where in contrast MPM can be quite dynamic. Those that do move, move at most to one neighboring grid-node or stay in the same node due to the CFL condition, chapter \ref{sec:cfl}. These methods often use fixed bin-sizes which can be quite memory intensive and add additional memory management. Fixed size binning make the methods mentioned before easier to implement which don't necessarily map well to the GPU due to their high complexity and uncoalesced accesses. As a result \cite{PIC:GPU} come to the conclusion to fall back to the CUDA library \texttt{thrust}, although they do note that the aforementioned methods can perform faster for particles that move only a limited amount of grid nodes. \texttt{thrust} provides the radix sort method \texttt{thrust::sort\_by\_key()}, where one sorts by grid index as key.

@@

This implementation compares index and deep counting sort and uses the latter. Although possible, this counting sort stays away from the limited movement implied by the CFL condition. Fixed size bins is an assumption this implementation although does not want to impose on the method.

\textbf{Particle-to-Grid-Transfers} are generally done in two manners \cite{stantchev2008fast}:
\begin{enumerate}
  \item \textbf{Particle Push:} X particles in a local neighborhood of a grid point push their attributes onto it. Since this relationship results in a race condition \texttt{atomicAdd(float f)}s or explicit synchronization is necessary, e.g. a\-tomic: \cite{PIC:GPU}, synchronized: \cite{FRANCESCO:ROSSI}.
  \item \textbf{Particle Pull:} A grid point pulls the particles in its local neighborhood from a sorted particle list. This has the benefit of no race conditions being present.
\end{enumerate}
A Grid-to-Particle-Transfer is in general easier since a particle already knows, due to its position, the grid points in its local neighborhood. The grid points in a uniform grid are 'sorted' by nature. This implementation tests one Particle Pull and two Particle Push methods.

Generally these methods are augmented by splitting the grid in 2D/3D blocks making use of the shared memory structure of the GPU \cite{PIC:GPU} \cite{honig2010generic} \cite{MPM:GPU}, often called domain decomposition. Shared memory is however limited per GPU (GTX 970, 48KB per block). The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a block. Since the number of particles per grid point in the MPM is generally between 4-10, only very limited block sizes are available for pull methods that greedily load all particles \cite{MPM:SNOW}.

@@

In Particle Push methods partial sums are also computed on the halo. Two different techniques exist to get the data between blocks coherent again:
\begin{enumerate}
\item Atomically add up the values of the halo. This results in two atomic adds on a side node, four on an edge node, and eight on a corner node; If the block extend is bigger than half the halo extend, otherwise more.
\item Every halo of every block corresponds to actual global memory. After a transfer, do add steps along each of the three axes. This is infeasible for large supports due to the added memory requirements. \cite{crassin2011interactive}
\end{enumerate}

\textbf{Particle Activation:} If the particle's velocity is under a user-defined threshold, it becomes deactivated. It does not get recognized by any Particle-to-Grid-Transfer, nor the grid solve(, nor any Grid-to-Particle-Transfer not affected by the velocity transfer). They may of course be reactivated if the velocity in the neighborhood exceeds that threshold again or they collide. \cite{MPM:GPU}

@@

\textbf{Particle Resampling:} A method that could alleviate pressure off fixed bin sizes and make shared-memory restrictions less of a factor is particle resampling. Their intention mostly is to fill numerical gaps in the material at large deformations where the material is not supposed to break yet. Split and merge methods can directly control how many particles in a cell are allowed and accordingly increase or reduce them.

@@

Unconditionally keeping particles on a fixed size is not desirable due to fracture requiring that gaps will be created. \cite{gao2017adaptive} produce a signed distance field to identify interior points. Surface points also should not merge or split due to visual artifacts. Furthermore, the signed distance field can be used to adapt the grid size depending on the distance to the surface. This allows for coarser levels with increasing distance from the surface.
%look again for MPM/PIC gpu impl
%SPH, FEM methods
\clearpage
\section{Notation}\label{sec:notation}
Multi-component types are generally printed in bold letters $\boldsymbol{A}$, $\boldsymbol{a}$. Vectors furthermore use lower-case letters $\boldsymbol{v}$ with the exception of angular momentum $\boldsymbol{L}$; Matrices use upper-case letters $\boldsymbol{M}$. A variable in the Lagrangian form has a left subscript 0: $ _0{\boldsymbol{x}}$. The Eulerian description has a left subscript t: $_t\boldsymbol{x}$. Occasionally notation may be omitted where it is apparent from context.

Beginning from chapter \ref{sec:mat_point} variables defined on a particle will have right subscript $p$ (e.g. $\boldsymbol{x}_p$). Grid cells will be assigned a right subscript $i$ or additionally $j$ denoting the grid index (e.g. $\boldsymbol{x}_i$). This is not to be confused with the following Einstein-Notation. Therefore beginning with chapter \ref{sec:mat_point} Einstein-Notation for components will use greek letters ($\alpha,\beta$). Occasionally summation over grid index $i$ and particles $p$ may be implied. A variable of the $n$-th time step with associated time $t^n = \sum_{i=1}^{n-1} \Delta t^i$ will have a right superscript: $\boldsymbol{x}^n$.
\subsection{Einstein-Notation}
Used throughout this thesis is at instances the Einstein-Notation when using tensors and vectors.

@@

\cite{MCGINTY:CONTINUUM}

@@

The first chapters will derive the Material Point Method and necessary extensions for an elastic model. An elastic model is an interesting case displaying much of the needed computations for a typical model. The last chapter \ref{sec:gpgpu} tends to the GPGPU side.

@@

Particles (or material points) in continuum mechanics are not what classically might be thought of as a particle. Rather the continuum assumption holds: each particle represents a continuous piece of material s.t. a microscopic view does not need to be adopted. A particular body is composed of a set of particles and can adapt different configurations due to changes in shape. These changes are caused by external or internal effects (forces etc.) on it and deform the body over time. \cite{MIT:CONTINUUM_MECHANICS} \cite{MPM:COURSE}

When modelling solids, changes of quantities from an initial reference configuration $_0\boldsymbol{x}$ to another current configuration $_t\boldsymbol{x}$ need to be measured. In the Material Point Method the reference configuration $_0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$). This is similar to the total Lagrangian formulation in finite element methods \cite{bathe2006finite}.

Let $\Omega ^ { 0 } , \Omega ^ { t } \subset \mathbb { R } ^ {  { 3 } }$ be the set of (material) points in the reference and current configuration respectively.  Then one may define a function or mapping ${^t_0}\phi ( \cdot , t ) : \Omega ^ { 0 } \rightarrow \Omega ^ { t }$, which relates the reference configuration to the current configuration. If for simplicity $_0\boldsymbol{x}$ and $_t\boldsymbol{{x}}$ describe the position of the particle in their respective configurations, this mapping becomes the deformation of the body from the reference configuration $_0\boldsymbol{x}$:
\begin{equation}
  _t\boldsymbol{x} =  {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } )
\end{equation}
If, for instance, the body consisting of each material point $_t\boldsymbol{x}$ moves with velocity $\boldsymbol{v}$ and rotation $\boldsymbol{R}(t)$ , this mapping is defined as:
\begin{equation}\label{eq:rigid}
  _t\boldsymbol{{x}} = {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } ) = \boldsymbol{R}(t)_0\boldsymbol{x}+\boldsymbol{v}t
\end{equation}
The velocity of a material point in the reference configuration can be defined using this mapping
\begin{equation}\label{eq:velocity}
  _0\boldsymbol{v}(_0\boldsymbol{x},t) = \frac{\partial ({^t_0}\phi)}{\partial t}(_0\boldsymbol{x},t),
\end{equation}
and similarly the acceleration is defined
\begin{equation}
  _0\boldsymbol{a}(_0\boldsymbol{x},t) = \frac{\partial^2 ({^t_0}\phi)}{\partial t^2}(_0\boldsymbol{x},t) = \frac{\partial _0\boldsymbol{v}}{\partial t}(_0\boldsymbol{x},t).
\end{equation}
It is helpful to abstract away from the reference configuration and think of it as being defined in a different fixed material space. Physically this has the impact of moving with the particle in world space. Commonly known as the Lagrangian form. It is often easier in continuum mechanics to start with a Lagrangian description and switch to a Eulerian one if needed. The Eulerian description is static: Variables of particles moving by are measured while staying in a fixed position.

These descriptions are different but they will yield the same measurements when related correctly. I.e. both configurations refer variables defined on them to the deformed state but the position, where the 'look up' of that value happens, is different. In the reference confguration look up is done at the initial position of the particle. In the current configuration look up happens at the particle's world position. These relations for some particle quantity $f$ are called the (Lagrangian) pull back
\begin{equation}
  _0f(_0\boldsymbol{x},t)= {_tf}({^t_0}\phi(_0\boldsymbol{x},t),t)
\end{equation}
and the (Eulerian) push forward
\begin{equation}\label{eq:push_forward}
  _tf(_t\boldsymbol{x},t) =  {_0f}({^t_0}\phi^{-1}(_t\boldsymbol{x},t),t) = {_0f}({_t^0}\phi(_t\boldsymbol{x},t),t)
\end{equation}
with definitions over their respective spaces $_tf ( \cdot , t ) : \Omega ^ {  { t } } \rightarrow \mathbb { R }$, $_0f ( \cdot , t ) : \Omega ^ {   { 0 } } \rightarrow \mathbb { R }$.  To enable the operator ${^t_0}\phi$ to be homeomorphic, s.t. an inverse ${^t_0}\phi^{-1} = {_t^0}\phi$ is defined, it is assumed that no two particles will ever occupy the same space at the same time.

@@

\subsection{Polar- and Singular Value Decomposition}\label{sec:svd}

@@

As can be seen the deformation gradient contains a rigid rotation. For strain measures this is not beneficial as an assumption of the stiffness tensor requires no net-rotation (more in chapter \ref{sec:linear_elasticty}). I.e. the deformation gradient has two components a constant rotation and the actual distortion or strain. There is two ways to deal with this:

@@

\subsubsection{Linear Elasticity}\label{sec:linear_elasticty}

@@

\subsubsection{Corotated Hyperelasticity}\label{sec:cor_hyper}

@@

\subsubsection{Fixed Corotated Hyperelasticity}

@@

Under material relaxation these stepping algorithms are attracted to a material-dependent primary contour. For instance for a model that aims for high volume preservation (Poisson ratio $\nu =0.5$) the primary contour is primarily enforcing volume-preservation ($ J_E = \sigma_1\sigma_2\sigma_3 \approx 1$). Reaching the rest configuration is only the secondary goal.

@@

The conservation of momentum (eq. \ref{eq:eul_transport}) can then be applied to the result of plugging \ref{eq:surface_ang_mom} back into \ref{eq:gen_ang_mom} leaving only:

@@

Before deriving the weak form an explanation of what it achieves is in order. The previous presented governing equations are written in the strong form: a solution to the equation needs to be exact on the whole domain and is as such influenced by the whole domain in general. Such a solution can be found for simplified models analytically. These act as a ground truth to numerical methods like the Material Point Method. Analytical methods of today can't handle complex problems: Numerical solutions try to overcome that hurdle.

@@

In the Material Point Method stress computations are more naturally done in the current configuration or equally in terms of the Cauchy stress as seen in chapter \ref{sec:linear_elasticty}. Pushing the equation forward to Eulerian view with the push forward $_t\boldsymbol{q}:\Omega^t \rightarrow \mathbb{R}^d$ of $_0\boldsymbol{q}$ is only a problem for the last term:

@@

\subsection{Material Point Method}\label{sec:mat_point}

>>>>

\noindent Countingsort für jede Partikel-Variable erhöht Coalescing und L2-Cache Hitraten. Binning teilt das Gitter in Blöcke auf, die Shared-Memory Implementationen er\-mög\-lich\-en. Die Größen der Bins sind für keine Operation beschränkt. Weiterhin werden nur Blöcke ausgeführt, die Informationen erhalten.

\vfill
\begin{center}
\subsubsection*{Abstract}
\end{center}
The material point method is allowing for physically based simulations. It has found its way into computer graphics and since then rapidly expanded. The material point method's hybrid use of Lagrangian particles as a persistent storage and a background uniform Eulerian grid enables solving of various partial differential equations with ease.

\noindent The material point method suffers from high execution times and is thus only viable for hero shots. The method is however highly parallelizable. Thus, this thesis proposes how to accelerate the material point method using GPGPU techniques.

\noindent Core of the material point method are grid and particles transfers that interpolate between the two structures. These transfers are executed multiple times per physical time step. Preprocessing steps might be taken if their computing time is outweighed.

\noindent Deep sorting with counting sort increases coalescing and L2 cache hit rates. Binning allows to divide the grid into blocks for shared memory filtering techniques. All operations do not rely on fixed bin size. As another preprocessing step, only grid blocks are executed which have particles in them.
\vfill
\clearpage
\tableofcontents

@@

\section{Introduction} \label{intro}
General purpose computation on graphics processing unit(GPGPU) has elevated graphics processing units(GPUs) to any computational field utilizing massively parallelizable algorithms. As such is the nature of discretizations in physics.

The higher parallelization acceleration of GPUs is a blessing for those applications that can profit from it. In unison with efficient algorithms they still have much potential to offer. Conversely, algorithms which are not parallelizable perform much worse and should be avoided. Dividing up these tasks between CPU and GPU is not recommendable. Transferring data over the PCI-Bus between the two is slow for big amounts of data and requires synchronization. Tasks in such a pipeline should thus stay on the GPU although they might perform worse.

@@

The simulation of the dynamics of materials is still extremely challenging in different applications and fields. Multiphysics interactions between materials are not realized on a big scale and often are not possible or feasible. The material point method(MPM) has already reliable tackled many different models, materials, and even interaction between materials.

Pure Eulerian methods like the finite element method(FEM) have been in use for long in engineering tasks but face problems handling topology changes with numerical stability. Pure Langrangian methods form the end other of the spectrum. Prominent are smoothed-particle hydrodynamics(SPH) that are restricted to (viscous) fluids and are of empirical nature. Particle in cell(PIC) techniques like the material point method are hybrid Lagrangian/Eulerian methods. They try to benefit from the nature of both. Advection on particles is trivial. Thus the material point method utilizes Lagrangian particles as a consistent storage of dynamic properties. The governing equations of the physical problem can be discretized easily to and solved on a uniform Eulerian grid. The grid is unmoving and does not store any information between physical time steps.

Transfers between the Lagrangian and Eulerian are of need to bring the two together. These transfers are the focus of particle in cell techniques. These transfers are realized due to interpolation. Thus they create (estimable) numerical error and lead to dissipation. Moreover grid nodes are mostly outnumbered by particles leading to more numerical error as modes of the single particles are not realized by the grid.

The $p$ particles to one node relationship of a particle-to-grid(P2G)-transfer or inversely one node to $p$ particles of a grid-to-particle(G2P)-transfer is the main problem when realizing these operations on the GPU. When faced with the static architecture of the GPU, this will inevitably lead to branch divergence of threads within a subgroup as some nodes are assigned more particles than others.

Chapter \ref{sec:rel_work} looks at the relevant literature of PIC-Methods specifically on the GPU as well as material point method extensions. In chapter \ref{sec:notation} notations which will be used throughout the thesis are laid out. The basics of chapter \ref{sec:basics} cover the derivation of an elastic material in the MPM to get an understanding of the prominent operations. Furthermore GPU intrinsics, metrics and optimization opportunities are discussed. Chapter \ref{sec:implementation} tends to the implementation. The evaluation of the implementation follows up in chapter \ref{sec:eval}. As a conclusion chapter \ref{sec:conclusion} gives a summarizing overview and tends to future work.

\clearpage
\section{Related Work}\label{sec:rel_work}
\textbf{Particle in cell} techniques were developed initially for hydrodynamics \cite{evans1957particle}. They are also widely used in plasma simulations to solve Maxwell's equations \cite{PIC:GPU}. Their key steps are identical to the MPM:
\begin{enumerate}
  \item Transfer particle data to a grid.
\item \label{it:solve_gov_eq}Solve the governing partial differential equations on the grid to move forward in time.
  \item Transfer grid data back to particles.
\end{enumerate}
In general the grid is regarded as a scratch-pad since its only a temporary storage for the means of computing item \ref{it:solve_gov_eq}. I.e. the grid could be deleted each physical times step. Based on PIC to reduce dampening and momentum conservation Fluid-Implicit-Particle(FLIP) was developed for fluid simulations and is still widely in use for large simulations. Their counterpart being the empirical smoothed-particle hydrodynamics \cite{gingold1977smoothed}.

As a further development the Affine-particle in cell(APIC) technique was introduced as an improvement to PIC and FLIP solvers. Due to the definition of a local affine velocity field around each particle (instead of a single vector) APIC conserves angular momentum, reduces damping and improves stability over PIC and FLIP, more in chapter \ref{sec:apic}. \cite{MPM:APIC} \cite{MPM:OLD_APIC}

@@

The \textbf{material point method} is a further development to the PIC methods. Both use analysis and numerics to be derived without the need for empirical assumptions. Originally it was developed by \cite{sulsky1995application} with a focus on the dynamics of solids. \cite{MPM:SNOW} brought the method into computer graphics for their animation film \textit{Frozen}. In rapid succession a variety materials and effects got modeled by the MPM. Since MPM is based on discretizing governing equations and using constitutive models covered by the different literature, its fast rate of expenditure is not a surprise.

@@

There exist several implementations utilizing the \textbf{GPU for particle in cell} techniques, which the material point method is a part of:

\textbf{Particle sorting:} \cite{PIC:GPU} studies different particle list sorting methods. Methods like 'Message Passing Sort' \cite{kong2011particle} \cite{decyk2011adaptable}, 'In Place Particle-QuickSort' \cite{stantchev2008fast} and 'Linked List Ordering' \cite{burau2010picongpu} try to utilize the partial sorting and only update particles that need to be moved within the list. These methods often rely on the assumption that only a small number of particles change their grid node, where in contrast MPM can be quite dynamic. Those that do move, move at most to one neighboring grid-node or stay in the same node due to the CFL condition, chapter \ref{sec:cfl}. These methods often use fixed bin-sizes which can be quite memory intensive and add additional memory management. Fixed size binning makes the methods mentioned before easier to implement which don't necessarily map well to the GPU due to their high complexity and uncoalesced accesses. As a result \cite{PIC:GPU} come to the conclusion to fall back to the CUDA library \texttt{thrust}, although they do note that the aforementioned methods can perform faster for particles that move only a limited amount of grid nodes. \texttt{thrust} provides the radix sort method \texttt{thrust::sort\_by\_key()}, where one sorts by grid index as key.

@@

This implementation compares index and deep counting sort and uses the latter. Although possible, this counting sort stays away from the limited movement implied by the Courant–Friedrichs–Lewy(CFL) condition. Fixed size bins are an assumption this implementation also does not want to impose on the method.

\textbf{Particle-to-grid-transfers} are generally done in two manners \cite{stantchev2008fast}:
\begin{enumerate}
  \item \textbf{Particle push:} X particles in a local neighborhood of a grid point push their attributes onto it. Since this relationship results in a race condition \texttt{atomicAdd(float f)}s or explicit synchronization is necessary, e.g. a\-tomic: \cite{PIC:GPU}, synchronized: \cite{FRANCESCO:ROSSI}.
  \item \textbf{Particle pull:} A grid point pulls the particles in its local neighborhood from a sorted particle list. This has the benefit of no race conditions being present.
\end{enumerate}
A \textbf{grid-to-particle-transfer} is in general easier since a particle already knows, due to its position, the grid points in its local neighborhood. The grid points in a uniform grid are 'sorted' by nature. This implementation tests one particle pull and two particle push methods.

Generally these methods are augmented by splitting the grid in 2D/3D blocks making use of the shared memory structure of the GPU \cite{PIC:GPU} \cite{honig2010generic} \cite{MPM:GPU}, often called domain decomposition. Shared memory is however limited per GPU (GTX 970, 48KB per block \cite{NVIDIA:GTX970}). The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a block. Since the number of particles per grid point in the MPM is generally between 4-10, only very limited block sizes are available for pull methods that greedily load all particles \cite{MPM:SNOW}.

@@

In particle push methods partial sums are also computed on the halo. Two different techniques exist to get the data between blocks coherent again:
\begin{enumerate}
\item Atomically add up the values of the halo. This results in two atomic adds on a side node, four on an edge node, and eight on a corner node; If the block extend is bigger than half the halo extend, otherwise more.
\item Every halo of every block corresponds to actual global memory. After a transfer, do add steps along each of the three axes. This is infeasible for large supports due to the added memory requirements. \cite{crassin2011interactive}
\end{enumerate}

\textbf{Particle activation:} If the particle's velocity is under a user-defined threshold, it becomes deactivated. It does not get recognized by any particle-to-grid-transfer, nor the grid solve(, nor any grid-to-particle-transfer not affected by the velocity transfer). They may of course be reactivated if the velocity in the neighborhood exceeds that threshold again or they collide. \cite{MPM:GPU}

@@

\textbf{Particle resampling:} A method that could alleviate pressure off fixed bin sizes and make shared-memory restrictions less of a factor is particle resampling. Their intention mostly is to fill numerical gaps in the material at large deformations where the material is not supposed to break yet. Split and merge methods can directly control how many particles in a cell are allowed and accordingly increase or reduce them.

@@

Unconditionally keeping particles on a fixed size is not desirable due to fracture requiring that gaps will be created. \cite{gao2017adaptive} produce a signed distance field to identify interior points. Surface points also should not merge or split due to visual artifacts. Furthermore, the signed distance field can be used to adapt the grid size depending on the distance to the surface. This allows for coarser levels with increasing distance from the surface.
%look again for MPM/PIC gpu impl
%SPH, FEM methods
\clearpage
\section{Notation}\label{sec:notation}
Multi-component types are generally printed in bold letters $\boldsymbol{A}$, $\boldsymbol{a}$. Vectors furthermore use lower-case letters $\boldsymbol{v}$ with the exception of angular momentum $\boldsymbol{L}$; Matrices use upper-case letters $\boldsymbol{M}$. A variable in the Lagrangian form has a left subscript $0$: $ _0{\boldsymbol{x}}$. The Eulerian description has a left subscript $t$: $_t\boldsymbol{x}$. Occasionally, notation may be omitted where it is apparent from context.

Beginning from chapter \ref{sec:mat_point} variables defined on a particle will have right subscript $p$ (e.g. $\boldsymbol{x}_p$). Grid cells will be assigned a right subscript $i$ or additionally $j$ denoting the grid index (e.g. $\boldsymbol{x}_i$). This is not to be confused with the following Einstein-notation. Therefore, beginning with chapter \ref{sec:mat_point} Einstein-notation for components will use greek letters ($\alpha,\beta$). Occasionally summation over grid index $i$ and particles $p$ may be implied. A variable of the $n$-th time step with associated time $t^n = \sum_{i=1}^{n-1} \Delta t^i$ will have a right superscript: $\boldsymbol{x}^n$.

Used throughout this thesis is at instances the \textbf{Einstein-notation} when using tensors and vectors.

@@

\begin{flushright}\cite{MCGINTY:CONTINUUM}\end{flushright}

@@

The first chapters will derive the material point method and necessary extensions for an elastic model. An elastic model is an interesting case displaying much of the needed computations for a typical model. The last chapter \ref{sec:gpgpu} tends to the GPGPU side.

@@

Particles (or material points) in continuum mechanics are not what classically might be thought of as a particle. Rather, the continuum assumption holds: Each particle represents a continuous piece of material s.t. a microscopic view does not need to be adopted. A particular body is composed of a set of particles and can adapt different configurations due to changes in shape. These changes are caused by external or internal effects (forces etc.) on it and deform the body over time. \cite{MIT:CONTINUUM_MECHANICS} \cite{MPM:COURSE}

When modelling solids, changes of quantities from an initial reference configuration $_0\boldsymbol{x}$ to another current configuration $_t\boldsymbol{x}$ need to be measured. In the material point method the reference configuration $_0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$). This is similar to the total Lagrangian formulation in finite element methods \cite{bathe2006finite}.

Let $\Omega ^ { 0 } , \Omega ^ { t } \subset \mathbb { R } ^ {  { 3 } }$ be the set of (material) points in the reference and current configuration respectively.  Then one may define a function or mapping ${^t_0}\phi ( \cdot , t ) : \Omega ^ { 0 } \rightarrow \Omega ^ { t }$ which relates the reference configuration to the current configuration. Let for simplicity $_0\boldsymbol{x}$ and $_t\boldsymbol{{x}}$ describe the position of the particle in their respective configurations. Then this mapping becomes the deformation of the body from the reference configuration $_0\boldsymbol{x}$:
\begin{equation}
  _t\boldsymbol{x} =  {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } )
\end{equation}
If for instance the body consisting of each material point $_t\boldsymbol{x}$ moves with velocity $\boldsymbol{v}$ and rotation $\boldsymbol{R}(t)$ , this mapping is defined as:
\begin{equation}\label{eq:rigid}
  _t\boldsymbol{{x}} = {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } ) = \boldsymbol{R}(t)_0\boldsymbol{x}+\boldsymbol{v}t
\end{equation}
The velocity of a material point in the reference configuration can be defined using this mapping
\begin{equation}\label{eq:velocity}
  _0\boldsymbol{v}(_0\boldsymbol{x},t) = \frac{\partial ({^t_0}\phi)}{\partial t}(_0\boldsymbol{x},t),
\end{equation}
and similarly the acceleration is defined
\begin{equation}
  _0\boldsymbol{a}(_0\boldsymbol{x},t) = \frac{\partial^2 ({^t_0}\phi)}{\partial t^2}(_0\boldsymbol{x},t) = \frac{\partial _0\boldsymbol{v}}{\partial t}(_0\boldsymbol{x},t).
\end{equation}
It is helpful to abstract away from the reference configuration and think of it as being defined in a different fixed material space. Physically this has the impact of moving with the particle in world space. Commonly known as the Lagrangian form. It is often easier in continuum mechanics to start with a Lagrangian description and switch to a Eulerian one if needed. The Eulerian description is static: Variables of particles moving by are measured while staying in a fixed position.

These descriptions are different but they will yield the same measurements when related correctly. I.e. both configurations refer variables defined on them to the deformed state but the position, where the 'lookup' of that value happens, is different. In the reference confguration lookup is done at the initial position of the particle. In the current configuration lookup happens at the particle's world position. These relations for some particle quantity $f$ are called the (Lagrangian) pull back
\begin{equation}
  _0f(_0\boldsymbol{x},t)= {_tf}({^t_0}\phi(_0\boldsymbol{x},t),t)
\end{equation}
and the (Eulerian) push forward
\begin{equation}\label{eq:push_forward}
  _tf(_t\boldsymbol{x},t) =  {_0f}({^t_0}\phi^{-1}(_t\boldsymbol{x},t),t) = {_0f}({_t^0}\phi(_t\boldsymbol{x},t),t)
\end{equation}
with definitions over their respective spaces $_tf ( \cdot , t ) : \Omega ^ {  { t } } \rightarrow \mathbb { R }$, $_0f ( \cdot , t ) : \Omega ^ {   { 0 } } \rightarrow \mathbb { R }$.  To enable the operator ${^t_0}\phi$ to be homeomorphic, s.t. an inverse ${^t_0}\phi^{-1} = {_t^0}\phi$ is defined, it is assumed that no two particles will ever occupy the same space at the same time.

@@

\subsection{Polar and singular value decomposition}\label{sec:svd}

@@

As can be seen, the deformation gradient contains a rigid rotation. For strain measures this is not beneficial as an assumption of the stiffness tensor requires no net-rotation (more in chapter \ref{sec:linear_elasticty}). I.e. the deformation gradient has two components a constant rotation and the actual distortion or strain. There are two ways to deal with this:

@@

\subsubsection{Linear elasticity}\label{sec:linear_elasticty}

@@

\subsubsection{Corotated hyperelasticity}\label{sec:cor_hyper}

@@

\subsubsection{Fixed corotated hyperelasticity}

@@

Under material relaxation these stepping algorithms are attracted to a material-dependent primary contour. For instance for a model that aims for high volume preservation (Poisson ratio $\nu =0.5$) the primary contour is primarily enforcing volume preservation ($ J_E = \sigma_1\sigma_2\sigma_3 \approx 1$). Reaching the rest configuration is only the secondary goal.

@@

The conservation of momentum (equation \ref{eq:eul_transport}) can then be applied to the result of plugging \ref{eq:surface_ang_mom} back into \ref{eq:gen_ang_mom} leaving only:

@@

Before deriving the weak form an explanation of what it achieves is in order. The previous presented governing equations are written in the strong form: a solution to the equation needs to be exact on the whole domain and is as such influenced by the whole domain in general. Such a solution can be found for simplified models analytically. These act as a ground truth to numerical methods like the material point method. Analytical methods of today can't handle complex problems: Numerical solutions try to overcome that hurdle.

@@

In the material point method stress computations are more naturally done in the current configuration or equally in terms of the Cauchy stress as seen in chapter \ref{sec:linear_elasticty}. Pushing the equation forward to Eulerian view with the push forward $_t\boldsymbol{q}:\Omega^t \rightarrow \mathbb{R}^d$ of $_0\boldsymbol{q}$ is only a problem for the last term:

@@

\subsection{Material point method}\label{sec:mat_point}
